---
title: 'Functional Enrichment Analysis'
author: |
 | Project: `r CIDAtools::ProjectName()`
 |
 | Analyst: `r CIDAtools::ProjectAnalyst()`
 |
 | Investigator(s): `r CIDAtools::ProjectPI()`
 |
 | Report generated: `r paste(format(Sys.Date(), '%B %d, %Y'))`
output:
  html_document:
    highlight: espresso
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
logo <- system.file("figures", "sph_cida_wm_blk.png", package="CIDAtools")
knitr::include_graphics(logo)
```

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(table1)
library(diffEnrich)
library(here)
```

# Project Summary

# Introduction

Introductory Text

# Methods

Functional Enrichment was run twice. Once using a list of the top 200 genes as determined by FDR-adjusted p-values, and once using a list of the top 200 genes as determined by FDR-adjusted p-values after sub-setting the results by |Log~2~(Fold-Change)| > 2. All analyses were performed using the `diffEnrich` package, which uses pathways defined by KEGG. For both of the gene lists mentioned, the background gene list was subset to include only the genes tested by the *Vape Status* model presented in previous reports. 

# Results 
```{r}
source(here("Code/07_Functional_Enrichment.R"))
source(here("Code/07a_Functional_Enrichment_no_cutoff.R"))


```

## Main Analysis

# Conclusions

Interpretations, discussion

# References

# Appendix

<!-- footer -->

---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
knitr::include_graphics(logo)
```

