---
title: Metadata Exploratory Report
author: |
 | Project: `r CIDAtools::ProjectName()`
 |
 | Analyst: `r CIDAtools::ProjectAnalyst()`
 |
 | Investigator(s): `r CIDAtools::ProjectPI()`
 |
 | Report generated: `r paste(format(Sys.Date(), '%B %d, %Y'))`
output:
  html_document:
    highlight: espresso
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
    code_folding: hide
---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
logo <- system.file("figures", "sph_cida_wm_blk.png", package="CIDAtools")
knitr::include_graphics(logo)
```

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(arsenal)
library(here)
library(readr)
library(coin)
library(latex2exp)
library(patchwork)
```

# Background

The data presented in this report are part of a study aimed to assess differential gene expression and methylation in vaping versus non-vaping LatinX youths in Pueblo and Denver, CO. Pulmonary function data were also obtained in order to better understand the impacts of vape use on pulmonary function. To assess differential gene expression and methylation, naso-epithelial swabs were obtained from each participating subject. Pulmonary function is assessed using PFTs (Pulmonary Function Tests) and Impulse Oscillometry (IOS).

# Study Population

  * This data set consists of samples taken from 51 people ages 12-17 from the Pueblo, Denver, and Aurora, CO areas.  
  * Vape Status (did you vape in the last 6 months?) and ethnicity are self-reported variables

# Methods
All analyses performed using `r R.version$version.string`
## Clinical Data Pocessing
  * Subjects are dichotomized to those that used a vaping device in the last 6 months and those who have not based on the variables 'ever_vape', 'vape_days', and 'last_vape'. *Six Month Vape Status* could not be confirmed for one individual. Any analyses including *Six Month Vape Status* will have a sample size of n = 50 assuming no other missing values.  
  Vaping status is self-reported. Previous analyses showed n = 12 participants had vaped in the last 6 months. This analysis will use n = 13 participants who had vaped in the last 6 months. One participant (SID = 111) reported that they had used a vaping device 5 out of the last 30 days, but did not respond to 'last_vape'. They were falsely labeled as "NA" in previous analyses.
  * Subjects' geographic location (City) is grouped by their reporting recruiting center. 
  
## Gene-Count Processing  
  * Library Size\:
    *
  * Normalization\:
  * Technical effects\:
  * Normalization\:

# Descriptive Statistics

*Table 1: Clinical Data*
```{r Table 1, message=FALSE, results='asis', echo=TRUE}
#import data
tab1_dat <- read_csv(here("DataProcessed/metadata_cleaning/table1_clean_data_2022_02_28.csv"))

tab1_dat$grade <-ordered(tab1_dat$grade,levels = c('7th', '8th','Freshman','Sophomore','Junior','Senior'))

#filter out 1 NA for vape status
tab1_dat <- tab1_dat %>% 
  filter(!is.na(vape_6mo_lab))

#create table 1
vape_6mo_table1 <- tab1_dat %>% 
  tableby(vape_6mo_lab ~ gender_lab + age + grade +
            recruitment_center + latino_lab + fev1_fvc + 
            r5 + x20, data = ., digits = 1)

#Fix Labels
labels(vape_6mo_table1) <- c(vape_6mo_lab = "Vaped in last 6 months", 
                             age = "Age", gender_lab = "Gender", recruitment_center = "City", 
                             latino_lab = "Ethnicity", grade = "Grade",
                             fev1_fvc = "FEV1/FVC", r5 = 'R5', x20 = 'X20')

#Print Tables
summary(vape_6mo_table1, pfootnote = T)
```

## Figures
```{r figures, echo=TRUE, message=FALSE, warning=FALSE}
#By Gender (n = 50)
bar_gender <- tab1_dat %>%   
  ggplot(aes(x = gender_lab, fill = vape_6mo_lab)) +
  geom_bar() +
  labs(x = "Gender", y = "Count") +
  ggtitle("Vaped in the last 6 months by Gender (n = 50)")

#By Latino(n = 50)
bar_latino <- tab1_dat %>% 
  ggplot(aes(x = latino_lab, fill = vape_6mo_lab))+
  geom_bar()+
  labs(x = "Ethnicity", y = "Count") +
  ggtitle("Vaped in the last 6 months by Latino (n = 50)")


#By FEV1/FVC Continuous
hist_fev1_fvc <- tab1_dat %>%
  ggplot(aes(x = fev1_fvc, fill = vape_6mo_lab))+
  geom_histogram(binwidth = 0.1, color = 'black', breaks = seq(0.5,1,0.1))+
  labs(x = TeX("\\frac{FEV1}{FVC}"), y = "Count") +
  xlim(0.5,1) +
  ggtitle(TeX("\\frac{FEV1}{FVC} \ by Vape Status (n = 28)"))

#By FEV1/FVC Box
box_fev1_fvc <- tab1_dat %>%
  ggplot(aes(x = vape_6mo_lab, y = fev1_fvc, fill = vape_6mo_lab)) +
  geom_boxplot(show.legend = F) +
  labs(x = "Vape Status", y = TeX("\\frac{FEV1}{FVC}")) +
  ggtitle("R5 by Vape Status (n = 28)")

#By R5(hist)
r5_hist <- tab1_dat %>%
  ggplot(aes(x = r5, fill = vape_6mo_lab)) +
  geom_histogram(binwidth = 1, color = 'black', breaks = seq(1,8,1)) +
  labs(x = TeX("\\frac{FEV1}{FVC}"), y = "Count") +
  xlim(1,8) +
  ggtitle("R5 by Vape Status (n = 49)")

#by R5(box)
r5_box <- tab1_dat %>%
  ggplot(aes(x = vape_6mo_lab, y = r5, fill = vape_6mo_lab)) +
  geom_boxplot(show.legend = F) +
  labs(x = "Vape Status", y = "R5") +
  ggtitle("R5 by Vape Status (n = 49)")

#by X20(box)
x20_box <- tab1_dat %>%
  ggplot(aes(x = vape_6mo_lab, y = x20, fill = vape_6mo_lab)) +
  geom_boxplot(show.legend = F) +
  labs(x = "Vape Status", y = "X20") +
  ggtitle("X20 by Vape Status (n = 44)")

#by x20(his)
x20_hist <- tab1_dat %>%
  ggplot(aes(x = x20, fill = vape_6mo_lab)) +
  geom_histogram(color = "Black", bins = 10, breaks = seq(-1.5,2.5,.5)) +
  labs(x = 'X20', y = "Count") +
  ggtitle("X20 by Vape Status (n = 44)")

bar_gender

bar_latino

hist_fev1_fvc

r5_box

x20_box




```



# Notes


# Questions  


  
      


<!-- footer -->

---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
knitr::include_graphics(logo)
```

