---
title: Metadata Exploratory Report
author: |
 | Project: `r CIDAtools::ProjectName()`
 |
 | Analyst: `r CIDAtools::ProjectAnalyst()`
 |
 | Investigator(s): `r CIDAtools::ProjectPI()`
 |
 | Report generated: `r paste(format(Sys.Date(), '%B %d, %Y'))`
output:
  html_document:
    highlight: espresso
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
    code_folding: hide
---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
logo <- system.file("figures", "sph_cida_wm_blk.png", package="CIDAtools")
knitr::include_graphics(logo)
```

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(arsenal)
library(here)
library(readr)
library(coin)
```

# Background

The data presented in this report are part of a study aimed to assess differential gene expression and methylation in vaping versus non-vaping LatinX youths in Pueblo and Denver, CO. Pulmonary function data were also obtained in order to better understand the impacts of vape use on pulmonary function. 

# Study Population

This data set consists of samples taken from 51 people ages 12-17 from the Pueblo, CO and the Denver-metro area. Subjects were asked to identify as 'Latino' or 'Non-latino'. 

# Methods

To assess differential gene expression and methylation, naso-epithelial swabs were obtained from each participating subject. Pulmonary function is assessed using PFTs (Pulmonary Function Tests).

# Descriptive Statistics

*Table 1: Clinical Data*
```{r Table 1, message=FALSE, results='asis', echo=TRUE}
#import data
table1_dat <- read_csv(here("DataProcessed/metadata_cleaning/table1_clean_data_2022_02_18.csv"))

table1_dat$grade <-ordered(table1_dat$grade,levels = c('7th', '8th','Freshman','Sophomore','Junior','Senior'))


#create table 1
vape_6mo_table1 <- tableby(includeNA(vape_6mo_lab) ~ sex_lab + age + grade +
                           city + latino_lab + 
                           ever_vape_lab + vape_30_lab + fev1 + fvc, 
                           data = table1_dat, digits = 1)

#Fix Labels
labels(vape_6mo_table1) <- c(vape_6mo_lab = "Vaped in last 6 months", 
                             age = "Age", sex_lab = "Sex", city = "City", 
                             latino_lab = "Race", grade = "Grade",
                             ever_vape_lab = "Have you vaped Before?", 
                             vape_30_lab = "Have you vaped in the last 30 days?", 
                             fev1 = "FEV1", fvc = "FVC")

#Print Tables
summary(vape_6mo_table1, pfootnote = T)
```

# Questions/Notes

  1. We are in the process of going through Cuining's old files and have yet to verify the status of the 'vaping' variables. We cannot yet confirm that 'vaped in the last 6 months', 'vaped in the last 30 days', or 'ever vaped' are 100% accurate.  
  2. What pulmonary function variables are typically the most helpful to include in a Table 1? I have included FEV1 and FVC as placeholders for now.  
  3. p-values presented in this table are not verified. I am unsure whether to potentially use different tests than the defaults. 

<!-- footer -->

---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
knitr::include_graphics(logo)
```

